---
title: "gghighlight"
author: "Daniel Schiff"
date: "`r format(Sys.Date(),'%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, messag=FALSE, warning=FALSE)
```

## gghighlight

When first learning R, I remember talking about using `gghighlight()` to showcase data. I became familiar with `ggplot()` early on in my R programming, and wanted a unique way to showcase outliers or focus  on particular aspects of graphs. This is when someone suggested `gghighlight()` to me.

Throughout this post we will look at how to use `gghighlight` in different setting and graphs. Our examples will include scatter plots, line plots and bar charts.

### Packages and Data

```{r}

require(tidyverse)
require(gghighlight)

# Fisher's iris data (comes with every R installation)
tibble(iris)

# A simulated data to show how proportion of 1's in a binomial random sample
# converges to the theoretical proportion of 1's, here 0.3

set.seed(5) # to ensure reproducibility of random numbers

output <- expand.grid(i = 1:100, n = 1:500) %>% # Creates data.frame with all possible combinations
  mutate(t = sample(0:1, nrow(.), replace=T, prob = c(0.7, 0.3))) %>% 
  group_by(i) %>% 
  mutate(f = cumsum(t),
         p = f/n) %>% 
  ungroup() %>% 
  mutate(i = factor(i))

tibble(output)

```


## Scatter Plot

Notice in the scatter plots below we can see how gghighlight works based on the commands we implement into the code chunk. In the first code chunk we want the color by Species if the `Species == "setosa"`. This is the condition we put into the `gghighlight()` function. 


```{r}

g <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point() +
  gghighlight(Species == "setosa", use_direct_label = FALSE) + 
  labs(title = "Iris", subtitle = "Sepal Width compared to Sepal Length")
g
```

Next, we want the colors to be related to the `Species != "setosa"` when species does not equal setosa. This is why in the second graphic we can see the two different colors in relation to the two other species in the iris data set. Using the `keep_scales = TRUE` argument we can notice the legend shows all the colors of the plot, but only those that are conditioned in the `gghighlight()`. 

```{r}
g1 <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point() +
  gghighlight(Species != "setosa", use_direct_label = FALSE, keep_scales = TRUE) + 
  labs(title = "Iris", subtitle = "Sepal Width compared to Sepal Length")
g1


```


## Line Plot

Line Plots have a lot of unique opportunities within gghighlight. In the first plot we will highlight the lines that have a percentage value that never falls below `.3`. 

```{r}
l <- ggplot(output, aes(x = n, y = p, color = i, group = i)) + 
  geom_line() +
  geom_hline(yintercept = 0.3, linetype=2) +
  gghighlight(min(p) > .3) + 
  labs(title = "Average value converges to 0.3 as sample size increases", x = "Sample Size", y = "Average value",
       color = "Line ID") + 
  scale_y_continuous(breaks = c(seq(0,1,0.25), 0.30)) +
  theme_classic() 
l
```

As we dive deeper into line plots we can make changes to make the plots unique. In this second plot, we notice that there are five lines highlighted. We limit  the number of highlighted lines using the option `max_highlight = 5L` (in case you're wondering, the `L` specifies that the number `5` is treated as an *integer* and not as a *float*. This is done to ensure we have the right times and prevents running into some potential issues). We do it for the highest average value of each simulated run. We also have `use_direct_label = FALSE`, which hides the label that is shown in the above graph, and moves it to the guide/legend. We also implement `unhighlighted_params = list(size = .5, color = NULL, alpha = .3))`. This argument changes the thickness of the line, the color of hidden lines and the transparency of those lines. 


```{r}

l1 <- ggplot(output, aes(x = n, y = p)) + 
  geom_line(aes(color = i), size = 1.25) +
  gghighlight(max(f), max_highlight = 5L, use_direct_label = FALSE, 
              unhighlighted_params = list(size = 1, color = NULL, alpha = .3)) +
  labs(title = "Average value converges to 0.3 as sample size increases", x = "Sample Size", y = "Average value",
       color = "Line ID") + 
  theme_classic()
l1

```

One fascinating thing is that we can make really cool examples with `facet_wrap()`. What `facet_wrap()` does is provide subplots based on different values of specified variables. Here, we're highlighting 4 particular specified lines, each in it's own subplot. 

```{r}

l2 <- ggplot(output, aes(x = n, y = p)) + 
  geom_line(aes(color = i), size = 1.1) + 
  gghighlight(i %in% c(1, 25, 50, 75), use_direct_label = F, 
              unhighlighted_params = list(size = .5, color = NULL, alpha = .3)) +
  labs(title = "Average value converges to 0.3 as sample size increases", x = "Sample Size", y = "Average value",
       color = "Line ID")  + 
  theme_classic() + 
  facet_wrap(~ i)
l2

```

## Histogram Chart

With histogram and bar charts, we can do some really exciting visuals. In the first plot we want to see the groupings that have samples larger than 1 standard deviation from the mean. As we see in the chart below we can see the highlighted values for the Totals that are outside of 1 standard deviation of the mean. 

```{r}
output2 <-  output %>%
  filter(n == 500) %>%
  mutate(i = as.numeric(i),
         ID_Group = case_when(i <= 25 ~ "Group 1 (1-25)",
                              i > 25 & i <= 50 ~ "Group 2 (26-50)",
                              i > 50 & i <= 75 ~ "Group 3 (51-75)",
                              i > 75 ~ "Group 4 (76-100)"))


h <- ggplot(output2, aes(x = f)) + 
  geom_histogram(bins = 9, aes(fill = ID_Group)) + 
  gghighlight(f >= mean(f) + sd(f) | f <= mean(f) - sd(f), use_direct_label = FALSE) +
  labs(title = "Histogram of Total", 
       subtitle = "From simulation",
       x = "Total",
       y = "Frequency",
       fill = "Grouping of i")
h
```


We can also do this for other analyses, such as seeing if there are any statistical outliers within a histogram. Notice how the condition is formatted to achieve the result of the graph. One principle for identifying statistical outliers is to consider points more than 1.5 $\times$ the *interquartile range* `IQR()` larger than the 3rd quartile or 1.5 $\times$ smaller than the 1st quartile to be outliers. This is what is implemented in the `geom_boxplot` and `boxplot` functions. 

> I believe the IQR computation is done group-wise rather than using all the data, since we're grouping the data first using `aes(fill = ID_group)`

```{r}

h1 <- ggplot(output2, aes(x = f)) + 
  geom_histogram(bins = 9, aes(fill = ID_Group)) + 
  gghighlight(f >= quantile(f, probs = .75) + 1.5*IQR(f) | 
                f <= quantile(f, probs = 0.25) - 1.5*IQR(f), 
use_direct_label = FALSE) +
  labs(title = "Histogram of Total", 
       subtitle = "From Simulation",
       x = "Total",
       y = "Frequency",
       fill = "Grouping of i")
h1

```

